---
title: "303 Project Proposal"
author: "Jasmine Tian"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("car")
#install.packages("rms")
#install.packages("gt")
#install.packages("kableExtra")

library(gt)
library(rms)
library(car)
library(readr)
library(kableExtra)
```

```{r}
data <- read.csv("default of credit card clients.csv")

# Clean the original data since he original data only defined 4 levels of education, 
# so clean out the other values
data <- data[data$X3 %in% c(1, 2, 3, 4), ]

# The original data defined 3 marital status: married, single, others
# Clean out others values first
data <- data[data$X4 %in% c(1, 2, 3), ]

# Calculate the ratio of 1 and 2 of X4 in the original data
ratio_1 <- sum(data$X4 == 1) / sum(data$X4 %in% c(1, 2))
ratio_2 <- 1 - ratio_1  # Since it's a binary choice, ratio_2 is the complement of ratio_1

# Change the marital status to the binary variable
# For each 3 in data$X4, assign 1 or 2 based on the calculated ratio
set.seed(1007935415)  # Set seed for reproducibility
data$X4 <- sapply(data$X4, function(x) {
  if (x == 3) {
    sample(c(1, 2), 1, prob = c(ratio_1, ratio_2))
  } else {
    x
  }
})

# Now data$X4 has "others" replaced with 1 or 2 based on the original ratio of 1 and 2

write_csv(data, "cleaned data.csv")

default <- data$Y

# Categorical Predictors
sex <- data$X2 # 1 = male & 2 = female
edu_level <- data$X3 # 4 education levels for the clients
marriage <- data$X4 # If the client married or not
status_sep <- data$X6 # Status of repayment

# Numerical Predictors
bal_lim <- data$X1 # Balance limit amount of the client
age <- data$X5 # Age of the client
bill_sep <- data$X12 # Amount of bill statement in September
paid_sep <- data$X18 # Amount paid of the bill in September
```

\newpage
## EDA tables and plots
```{r}

# Assuming 'data' is your dataframe and 'bal_lim' is the numerical variable

# Compute the summary statistics
bal_lim_stats <- summary(bal_lim)

# Convert the stats to a data frame for nice formatting
bal_lim_df <- data.frame(
  Statistic = c("Min", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max"),
  Value = as.numeric(bal_lim_stats)
)

# Create a nicely formatted table
kable(bal_lim_df, caption = "Summary Statistics for Balance Limit", 
      col.names = c("Statistic", "Value"), digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center") %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "2cm")
```

```{r}
library(ggplot2)
library(dplyr)
library(gridExtra)
# Count Plot of Marriage Status
p2 <- ggplot(data, aes(x = factor(marriage))) + 
  geom_bar(fill = "cornflowerblue") +
  labs(title = "Marriage Status Counts", 
       x = "Marriage Status", y = "Count", 
       caption = "Count of Marriage Status") +
  theme(plot.caption = element_text(hjust = 0.5)) +
  scale_x_discrete(labels = c("1" = "Married", "2" = "Single", "3" = "Others"))
#p2

# Count Plot of Count of Repayment Status in September, 2005
p3 <- ggplot(data, aes(x = factor(status_sep))) + 
  geom_bar(fill = "deepskyblue") +
  labs(title = "Count of Repayment Status in Sep", 
       x = "Repayment Status", y = "Count", 
       caption = "
       Repayment Status of Client in Sep, 2005") +
  theme(plot.caption = element_text(hjust = 0.5))
#p3

# Histogram of Bill Statement in September
p4 <- ggplot(data, aes(x = bill_sep)) + 
  geom_histogram(bins = 30, fill = "cyan3", color = "black") +
  labs(title = "Histogram of Bill Statement in Sep", 
       x = "Bill Statement", y = "Frequency", 
       caption = "Histogram of Bill Statement of Client in Sep, 2005") +
  theme(plot.caption = element_text(hjust = 0.5))
#p4

# Histogram of Amount Paid in September
p5 <- ggplot(data, aes(x = paid_sep)) + 
  geom_histogram(bins = 30, fill = "cyan4", color = "black") +
  labs(title = "Amount Paid in September", 
       x = "Amount Paid", y = "Frequency", 
       caption = "Histogram of Amount Paid of Client in Sep, 2005") +
  theme(plot.caption = element_text(hjust = 0.5))
#p5

grid.arrange(p2,p3,p4,p5, ncol = 2)
```

\newpage
## Modelling

```{r}
model1 <- glm(default~bal_lim + age + bill_sep + paid_sep + 
               sex + edu_level + marriage + status_sep, 
             family = binomial(link = logit))
summary(model1)
cat("BIC =", BIC(model1))
```
```{r}
## Stepwise elimination for model1 based on AIC ##
sel.var.aic <- step(model1, trace = 0, k = 2, direction = "both") 
select_var_aic<-attr(terms(sel.var.aic), "term.labels")   
select_var_aic
```

```{r}
## Stepwise elimination for model1 based on AIC ##
sel.var.bic <- step(model1, trace = 0, k = log(nrow(data)), direction = "both")
select_var_bic<-attr(terms(sel.var.bic), "term.labels")   
select_var_bic
```


```{r}
# Model 2 eliminates edu_level from model 1 since it has the least significant p-value in model 1, and sex since it's a categorical value
model2 <- glm(default~bal_lim + age + bill_sep + paid_sep + 
               marriage + status_sep, 
             family = binomial(link = logit))
summary(model2)
cat("BIC =", BIC(model2))
```

```{r}
## Stepwise elimination for model2 based on AIC ##
sel.var.aic <- step(model2, trace = 0, k = 2, direction = "both") 
select_var_aic<-attr(terms(sel.var.aic), "term.labels")   
select_var_aic
```

```{r}
## Stepwise elimination for model2 based on AIC ##
sel.var.bic <- step(model2, trace = 0, k = log(nrow(data)), direction = "both")
select_var_bic<-attr(terms(sel.var.bic), "term.labels")   
select_var_bic
```


```{r}
# Model 3 eliminates age & edu_level from model 1 since they have the least significant p-value in model 1
model3 <- glm(default~bal_lim + bill_sep + paid_sep + 
               sex + marriage + status_sep, 
             family = binomial(link = logit))
summary(model3)
cat("BIC =", BIC(model3))
```


```{r}
## Stepwise elimination for model3 based on AIC ##
sel.var.aic <- step(model3, trace = 0, k = 2, direction = "both") 
select_var_aic<-attr(terms(sel.var.aic), "term.labels")   
select_var_aic
```

```{r}
## Stepwise elimination for model3 based on AIC ##
sel.var.bic <- step(model3, trace = 0, k = log(nrow(data)), direction = "both")
select_var_bic<-attr(terms(sel.var.bic), "term.labels")   
select_var_bic
```
The third model has 6 predictors in total, with AIC of 27893 and BIC of 27951.43. Now we need to compare all three models and select the best mode.


### Model Selection

```{r}
aic_values <- c(AIC(model1), AIC(model2), AIC(model3))
bic_values <- c(BIC(model1), BIC(model2), BIC(model3))

# Create a data frame
model_comparison <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3"),
  AIC = aic_values,
  BIC = bic_values
)
```


```{r}
gt(model_comparison) %>%
  tab_header(
    title = "Model Comparison"
  ) %>%
  tab_source_note(
    source_note = "Table 1: Model Comparison using AIC and BIC"
  ) %>%
  cols_label(
    Model = "Model",
    AIC = "AIC",
    BIC = "BIC"
  ) %>%
  tab_options(
    column_labels.font.size = px(12),
    heading.title.font.size = px(14),
    source_notes.font.size = px(10)
  )

```
We choose model 3 as our final model.



\newpage

## Model calibration with cross-validation and bootstrap  for model3

```{r, eval=TRUE, echo = T}
## Fit the model with lrm from rms package ##
lrm.final <- lrm(model3, data = data, x = TRUE, y = TRUE)
cross.calib <- calibrate(lrm.final, method="crossvalidation", B=10) # model calibration
plot(cross.calib, las=1, xlab = "Predicted Probability for model3")
```


## Discrimination with ROC curve
```{r, eval=TRUE, echo = T}
library(pROC)
p <- predict(lrm.final, type = "fitted")
roc_logit <- roc(response = data$Y, predictor = p)
## The True Positive Rate ##
TPR <- roc_logit$sensitivities
## The False Positive Rate ##
FPR <- 1 - roc_logit$specificities

plot(FPR, TPR, xlim = c(0,1), ylim = c(0,1), type = 'l', lty = 1, lwd = 2,col = 'red')
abline(a = 0, b = 1, lty = 2, col = 'blue')
text(0.7,0.4,label = paste("AUC = ", round(auc(roc_logit),2)))

auc(roc_logit)
```


```{r}
vif(model3)
```


## Plot the dfbetas and deviance residuals

## Dfbetas
```{r, eval=TRUE, echo = T}
## Dfbetas ##

log.mod.final <- glm(default~bal_lim + bill_sep + paid_sep + 
               sex + marriage + status_sep, 
             family = binomial(link = logit))
df.final <- dfbetas(log.mod.final)
head(df.final)

par(family = 'serif')
par(mfrow = c(1, 3))
plot(data$X1, df.final[,1], xlab='Amount of Balance Limit (NT dollar)', 
     ylab='dfbeta')
lines(lowess(data$X1, df.final[,1]), lwd=2, col='blue')
abline(h=0, lty='dotted')
abline(h=-2/sqrt(nrow(df.final)), lty='dotted')
abline(h=2/sqrt(nrow(df.final)), lty='dotted')

plot(data$X12, df.final[,2], xlab='Amount of bill statement (NT
dollar) in September, 2005', 
     ylab='dfbeta')
lines(lowess(data$X12, df.final[,1]), lwd=2, col='blue')
abline(h=0, lty='dotted')
abline(h=-2/sqrt(nrow(df.final)), lty='dotted')
abline(h=2/sqrt(nrow(df.final)), lty='dotted')

plot(data$X18, df.final[,2], xlab='Amount paid (NT dollar) in
September, 2005', 
     ylab='dfbeta')
lines(lowess(data$X18, df.final[,1]), lwd=2, col='blue')
abline(h=0, lty='dotted')
abline(h=-2/sqrt(nrow(df.final)), lty='dotted')
abline(h=2/sqrt(nrow(df.final)), lty='dotted')
```

## Deviance residuals
```{r, eval=TRUE, echo = T}
## Plot the deviance residuals ##
res.dev <- residuals(log.mod.final, type = "deviance")
par(family = 'serif')
par(mfrow = c(1, 3))

plot(data$X1, res.dev, xlab='Amount of Balance Limit (NT dollar)', 
     ylab='Deviance Residuals')
lines(lowess(data$X1, res.dev), lwd=2, col='blue')
abline(h=0, lty='dotted')

plot(data$X12, res.dev, xlab='Amount of bill statement (NT dollar) in
September, 2005', 
     ylab='Deviance Residuals')
lines(lowess(data$X12, res.dev), lwd=2, col='blue')
abline(h=0, lty='dotted')

plot(data$X18, res.dev, xlab='Amount paid (NT dollar) in
September, 2005', 
     ylab='Deviance Residuals')
lines(lowess(data$X18, res.dev), lwd=2, col='blue')
abline(h=0, lty='dotted')
```
